---
title: "Working with PdfDocument objects"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(pdftextclusteR)
```

`pdf_read()` returns a `PdfDocument`: a list-like object with one
`PdfPage` per page. After `pdf_detect_clusters()` the pages are
`PdfClusters` objects — the same pages, with every word assigned to a
cluster. This article walks through the generics that work on all three.
The bundled `cibap` dataset is a `PdfDocument`, so we use it directly.

## Printing

Every object prints a one-line summary:

```{r}
cibap
cibap[[18]]

clusters <- cibap[[18]] |> pdf_detect_clusters(verbose = FALSE)
clusters
```

A clustered document also reports the algorithm:

```{r}
doc_clusters <- cibap[15:20] |> pdf_detect_clusters(verbose = FALSE)
doc_clusters
```

## Subsetting: `length()`, `[[` and `[`

A document behaves like a list of pages: `[[` returns one page, `[`
returns a smaller `PdfDocument`:

```{r}
length(cibap)
cibap[[18]]@number
cibap[15:20]
```

Note that pages keep their real page number: `cibap[15:20][[1]]` is page
15, and the `page` column in extracted text says 15, not 1.

## Properties: `@`

Object metadata lives in properties, accessed with `@`:

```{r}
cibap@source
cibap[[18]]@number
cibap[[18]]@width
cibap[[18]]@height
```

A `PdfClusters` page adds the algorithm and the parameters you passed:

```{r}
tuned <- cibap[[18]] |> pdf_detect_clusters(eps = 10, verbose = FALSE)
tuned@algorithm
tuned@params
```

`S7::props(x)` lists all properties at once.

## To plain data: `as_tibble()` and `as.data.frame()`

The word data behind any object is available as a plain tibble or data
frame. For a document, a `page` column is added:

```{r}
tibble::as_tibble(cibap[[18]]) |> head(3)
tibble::as_tibble(cibap[15:16]) |> head(3)
```

This is the escape hatch to dplyr and friends; the objects themselves
stay out of your data wrangling.

## Summaries: `summary()`

Each level summarizes what is useful at that level. A document: words
(and, once clustered, clusters) per page — the returned tibble is handy
for spotting problem pages:

```{r}
overview <- summary(doc_clusters)
dplyr::filter(overview, clusters > 15)
```

A page: the fonts in use — a quick feel for the structure:

```{r}
summary(cibap[[18]])
```

A clustered page: a per-cluster overview with a text preview:

```{r}
summary(clusters) |> head(5)
```

## Plotting: `pdf_plot_clusters()`, `plot()` and `autoplot()`

`pdf_plot_clusters()` is the documented entry point; `plot()` and
`ggplot2::autoplot()` do the same thing:

```{r}
plot(clusters)
```

All plotting options (`show_order`, `color_by`) work through every
entry point; see `?pdf_plot_clusters` and the article
[Text types and reading order](text-types.html).
