---
title: "Text types and reading order"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(pdftextclusteR)
```

## The pipeline

`pdftextclusteR` processes a document in four steps:

```r
doc <- pdf_read("document.pdf")     # 1. read: words + fonts + page sizes
clusters <- pdf_detect_clusters(doc)   # 2. cluster words into text blocks
classified <- pdf_classify_clusters(clusters)  # 3. assign text types
text <- pdf_extract_clusters(classified)       # 4. extract the text
```

The steps are deliberately separate:

* **Detection is per page and you typically run it more than once.** The
  usual workflow is a tuning loop: detect, plot, adjust parameters,
  detect again. Only the final result needs classification.
* **Classification is document-level.** Its strongest signals span pages:
  a footer is recognized because (nearly) the same text recurs at the
  same position on most pages, and a page number because its value
  increases in step with the page index. While clustering a single page
  that information does not exist yet, so the steps are sequential by
  nature.

When you do not need to inspect or tune anything, `pdf_extract_text()`
runs all four steps with sensible defaults — it also accepts a file path:

```{r}
text <- pdf_extract_text(cibap[15:20], verbose = FALSE)
head(text, 5)
```

## What the classifier looks at

`pdf_classify_clusters()` assigns one of these types to every cluster:

| Type | Meaning | Main signal |
|------|---------|-------------|
| `body` | running text | the default when nothing else matches |
| `heading` | section titles, with `.type_level` | font size relative to the body text, bold |
| `caption` | figure/table captions | text pattern ("Figuur", "Table", ...) |
| `figure_text` | chart and figure labels | few words, numeric, small font |
| `page_header` | running page header | repetition in the top band |
| `page_footer` | running page footer | repetition in the bottom band |
| `page_number` | standalone page numbers | numeric value tracks the page index |

Repetition is checked on *digit-masked* text, so "Rapport 2023 — pag. 4"
and "Rapport 2023 — pag. 5" count as the same footer.

```{r}
classified <- cibap |>
  pdf_detect_clusters(verbose = FALSE) |>
  pdf_classify_clusters()
```

`color_by = ".type"` shows the result. Page 18 is an infographic: the
title and chart titles are headings, the chart labels are `figure_text`,
and the footer is recognized:

```{r}
classified[[18]] |>
  pdf_plot_clusters(color_by = ".type")
```

## Text types and the reading order

Clusters are numbered in reading order (a recursive XY-cut over the
cluster bounding boxes; see `?pdf_detect_clusters`). Classification
improves that order: page furniture — headers, footers and page
numbers — is taken out of the XY-cut and numbered *after* the reading
flow, so the flow contains only real content.

`show_order = TRUE` draws the order as arrows:

```{r}
classified[[44]] |>
  pdf_plot_clusters(show_order = TRUE, color_by = ".type")
```

## Excluding types from the extraction

After classification the extraction carries `.type` and `.type_level`
columns, and `exclude` drops unwanted types:

```{r}
classified[[18]] |>
  pdf_extract_clusters(exclude = c("page_footer", "figure_text")) |>
  head(5)
```

`pdf_extract_text()` excludes `page_header`, `page_footer` and
`page_number` by default.

## Tuning the rules

All thresholds live in a rules object. Pass a modified copy to
`pdf_classify_clusters()`:

By default the margins are adaptive (`margins = "auto"`): headers and
footers are searched in wide bands (top and bottom 30% of the page), and
a repeated text only qualifies when it sits at a *stable vertical
position* across pages. This adapts to each document's actual margins —
important when you process many documents with different layouts. With
`margins = "fixed"` the bands are exactly `top_margin`/`bottom_margin`
and no stability is required.

```{r}
# Force narrow, fixed bands instead of the adaptive default
rules <- pdf_type_rules(margins = "fixed", top_margin = 0.1)

cibap[15:20] |>
  pdf_detect_clusters(verbose = FALSE) |>
  pdf_classify_clusters(rules = rules, verbose = FALSE)
```

See `?pdf_type_rules` for all thresholds. Two things to keep in mind:

* Repetition-based types need multiple pages (by default at least 3
  matching pages and 30% of the document). On a single page or a small
  subset, headers and footers will not be detected.
* When precision matters more than recall — excluded text is *gone* —
  prefer conservative thresholds: a missed footer is a small nuisance,
  body text wrongly excluded is data loss.
